### 2024-07-02 thesis/ms-fletcher-porter.tex
first draft

A ubiquitous example of a tech stack in IT is the \textit{web stack}, used for developing websites and related pieces of software. The task that websites must perform, as suggested by this stack, is to present an interface to a user through a web browser (the \textit{front end}), to conduct some business logic like creating a customized page for the user (the \textit{back end}), to perform network communication (the \textit{server}), and to store and dispatch data (the \textit{database})~\cite{PranamStack2017}. This doesn't describe all websites---many don't have databases or any real back-end logic, but it describes at least a superset of the functionality of most websites.

@@

In discussing the web stack, it's also useful to think about a particular set of tools used for a single project, i.e.\ one tool from each layer. Alternately, this can be thought of as the project's cross section of the web stack, or the project's own stack. There are several of these cross sections that are widely used, such as LAMP (Linux, Apache, MySQL, PHP) or MEAN (MongoDB, Express.js, Angular.js, Node.js)~\cite{PranamStack2017}. Just as often, firms employ stacks tailored to their technical and personnel needs and often write about these decisions in blog posts, such as for recent changes to Microsoft Teams in~\cite{Singh2023}.

@@

The model itself consists of layered abstractions over physical communication media, as seen in figure~\ref{fig:osi_model}. The lowest layer, Physical, concerns itself with the physical phenomena of communication. In the case of CAN, this means voltage levels, terminating resistors, twisted pairs, grounding rules, and more. The highest level, Application, provides network facilities without need for understanding \textit{how} the network connection is made. The Application layer includes, for example, identifying communication partners and negotiating security protocols. Well-known protocols like HTTP, DNS, and FTP exist in the Application layer. The intermediate layers provide increasing levels of abstraction over the physical layer, defining things like the structure of data packets, how to route messages, how to route them efficiently, and more.

@@

An interesting property of layered abstraction architectures is that protocols which are defined in higher layers can have the lower layers transparently replaced. For instance, the internet protocol which we use every day is defined in the Network and Transport layers. Typically, the internet is built over Ethernet or Wi-Fi, but with some effort, other low-level protocols are available. Lindgren et al.\ developed IP over CAN, wherein an internet network is extended over a CAN network~\cite{Lindgren2008}, which was never imagined by implementers of CAN nor IP.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=0.8\textwidth]{assets/dolores}
  \caption{Dolores rotates the top platform about a pivot by actuating a pair of hydraulic cylinders.  A pair of additional cylinders in parallel can resist the motion providing loads useful for testing.}\label{fig:dolores0}
\end{figure}

@@

The Solid Mechanics Laboratory at Aalto University conducts experiments to identify the properties of engineering materials under load particularly studying phenomena ranging from the micron to the meter scale.  As is typical for laboratories studying solid mechanics, they do this by subjecting test samples to a load schedule using a servo-controlled hydraulic cylinder.  They have several test beds, each with an instance of a common control system.  This had been controlled by a workstation-commanded DAQ that communicated with a motor controller and a strain gauge amplifier, but new experiments and the desire for faster, better research cycles called for improvements to the control system.

@@

An FPGA, like a CPU, is a programmable chip that performs computations.  However, they differ in when the program is read.  In a CPU, a program is read and executed as the program runs.  It reads an instruction from an address in memory specified by an instruction counter internal to the CPU.\  That instruction will be loaded into the CPU and in executing it, the CPU will change its state, perhaps by loading a value from memory into a CPU register, executing an arithmetic or logic instruction, interfacing with IO, or any of myriad other things.  An FPGA, on the other hand, reads and loads the entire program before execution begins.  It does this by mutating its internal hardware circuits to implement the program.  With this lower-level program implementation an FPGA is able to see higher performance for similar programs.  Instead of indirection through memory lookups, instruction counters, bus loads, and registers, an FPGA can pipeline all necessary instructions in a row.  A PI controller can be implemented as a subtraction circuit, followed by two gains and an accumulator circuit, followed by an addition circuit.  The circuit would be essentially passive in this case, though there can be memory for more complex logic like would be needed for a PD controller.  The programs for FPGAs are generally written in hardware description languages (HDLs) like Verilog or VHDL which have important hardware concepts like time core to the language, though there are libraries for conventional languages like C++ or Rust to generate HDL code.

@@

The control layer was built, like the lower layers, with NI tooling.  All of it was built with LabView, with the CPU-based controls using the LabView Real-Time Module~\cite{NIRealTimeModule} and the FPGA controls built with the LabView FPGA Module~\cite{NIFPGAModule}.  It implemented a PID controller, though the derivative term was in practice usually omitted.  To monitor the condition of long-running tests and for data acquisition, there are programs running on a desktop workstation that observe the state of the control system.  They also developed a custom scripting language that runs on the workstation to define the high-level testing schedule, separate from the low-level control logic.

The Aalto Solid Mechanics Laboratory's new control system

@@

This paper presented three case studies to demonstrate the utility of the novel theory of a mechatronic tech stack. These cases were diverse in their applications and methods and by viewing them through the lens of the tech stack, new knowledge of how fluid power systems can be commissioned was gained. Also, by studying several stacks, I have formed a pool of tools to draw from. The mechatronic tech stack can be re-rendered to reflect this, as here in figure~\ref{fig:mechatronic_tech_stack_filled}.

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{assets/mechatronic_tech_stack_filled}
  \caption{Not only does the mechatronic tech stack present layers of abstraction, it also provides an idea of what tools to choose, similar to figure~\ref{fig:web_stack}.}\label{fig:mechatronic_tech_stack_filled}
\end{figure}

@@

>>>>

A ubiquitous example of a tech stack in IT is the \textit{web stack}, used for developing websites and related pieces of software. The task that websites must perform, as suggested by this stack, is to present an interface to a user through a web browser (the \textit{front end}), to conduct some business logic like creating a customized page for the user (the \textit{back end}), to perform network communication (the \textit{server}), and to store and dispatch user data (the \textit{database})~\cite{PranamStack2017}. This doesn't describe all websites---many don't have databases or any real back-end logic, but it describes at least a superset of the functionality of most websites.

@@

In discussing the web stack, it's also useful to think about a particular set of tools used for a single project, i.e.\ one tool from each layer. Alternately, this can be thought of as the project's cross section of the web stack, or the project's own stack. There are several of these cross sections that are widely used, such as LAMP (Linux, Apache, MySQL, PHP) or MEAN (MongoDB, Express.js, Angular.js, Node.js)~\cite{PranamStack2017}. Just as often, firms employ stacks tailored to their technical and personnel needs and often write about these decisions in blog posts, such as for changes in 2023 to Microsoft Teams in~\cite{Singh2023}.

@@

The model itself consists of layered abstractions over physical communication media, as seen in figure~\ref{fig:osi_model}. The lowest OSI layer, Physical, concerns itself with the physical phenomena of communication. In the case of CAN, this means specifying voltage levels, terminating resistors, twisted pairs, grounding rules, and more. The highest level, Application, provides networking facilities without need for understanding \textit{how} the network connection is made. The Application layer includes, for example, identifying communication partners and negotiating security protocols. Well-known protocols like HTTP, SSH, and FTP exist in the Application layer. The intermediate layers provide increasing levels of abstraction over the physical layer, defining things like the structure of data packets, how to route messages, how to route them efficiently, and more.

@@

An important property of layered abstraction architectures is that protocols which are defined in higher layers can have the lower layers transparently replaced. For instance, the internet protocol (IP) which we use every day is defined in the Network layer. Typically, the internet is built over Ethernet or Wi-Fi, but with some effort, other low-level protocols are available. Lindgren et al.\ developed IP over CAN, wherein an internet network is extended over a CAN network~\cite{Lindgren2008}, which was never imagined by implementers of CAN nor IP.

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.8\textwidth]{assets/dolores}
  \caption{Dolores rotates the top platform about a pivot by actuating a pair of hydraulic cylinders.  A pair of additional cylinders in parallel can resist the motion providing loads useful for testing.}\label{fig:dolores0}
\end{figure}

@@

The Solid Mechanics Laboratory at Aalto University conducts experiments to identify the properties of engineering materials under load particularly studying phenomena ranging from the micron to the meter scale.  As is typical for laboratories studying solid mechanics, they do this by subjecting test samples to a load schedule using a servo-controlled hydraulic cylinder as seen in figure~\ref{fig:solid_mechanics_test_rig}.  They have several test beds, each with an instance of a common control system.  This had been controlled by a workstation-commanded DAQ that communicated with a motor controller and a strain gauge amplifier, but new experiments and the desire for faster, better research cycles called for improvements to the control system.  The following discussion is based heavily on an interview I took with the lab's head Dr.~Pauli Lehto.

\begin{figure}[h]
  \centering
  \includegraphics[width=0.90\textwidth]{assets/solid_mechanics_test_rig}
  \caption{A typical test in the solid mechanics lab, this has a sample will experience axial loads provided by the surrounding machine and sensors on and around the sample collect measurements for data acquisition and control feedback.}\label{fig:solid_mechanics_test_rig}
\end{figure}

@@

An FPGA, like a CPU, is a programmable chip that performs computations.  However, they differ in when the program is read.  In a CPU, a program is read and executed as the program runs.  It reads an instruction from an address in memory specified by an instruction counter internal to the CPU~\cite{Malvino1999}.  That instruction will be loaded into the CPU and in executing it, the CPU will change its state, perhaps by loading a value from memory into a CPU register, executing an arithmetic or logic instruction, interfacing with IO, or any of myriad other things.  An FPGA, on the other hand, reads and loads the entire program before execution begins~\cite{MeyerBaese2014a,Kaur2009}.  It does this by mutating its internal hardware circuits to implement the program.  With this lower-level program implementation an FPGA is able to see higher performance for similar programs.  Instead of indirection through memory lookups, instruction counters, bus loads, and registers, an FPGA can pipeline all necessary instructions in a row.  A PI controller can be implemented as a subtraction circuit, followed by two gains and an accumulator circuit, followed by an addition circuit.  The circuit would be essentially passive in this case, though there can be memory for more complex logic like would be needed for a PD controller.  The programs for FPGAs are generally written in hardware description languages (HDLs) like Verilog~\cite{IEEE1800-2023} or VHDL~\cite{IEEE1070-2000} which have important hardware concepts like time core to the language, though there are libraries for conventional languages like C++~\cite{VivadoHLS} or Rust~\cite{RustHDL} to generate HDL code.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=0.98\textwidth]{assets/solid_mechanics_control_monitor}
  \caption{The rack-mountable box on the right contains the analog IO components, the FPGA and CPU, and other supporting electronics.  It's connected to the desktop workstation on the left which can monitor the state of the system, provide an interface for operators to set test schedules, and record data acquisition.  Not pictured is the QuantumX data acquisition and amplifier unit.}\label{fig:solid_mechanics_control_monitor}
\end{figure}

The control layer was built, like the lower layers, with NI tooling.  All of it was built with LabView, with the CPU-based controls using the LabView Real-Time Module~\cite{NIRealTimeModule} and the FPGA controls built with the LabView FPGA Module~\cite{NIFPGAModule}.  It implemented a PID controller, though the derivative term was in practice usually omitted~\cite{Schad2011,Ali2010}.  To monitor the condition of long-running tests and for data acquisition, there are programs running on a desktop workstation, seen in figure~\ref{fig:solid_mechanics_control_monitor}, that observe the state of the control system.  They also developed a custom scripting language that runs on the workstation to define the high-level testing schedule, separate from the low-level control logic.

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{assets/mechatronic_tech_stack_solid_mechanics}
  \caption{The tech stack is dominated by National Instruments CompactRIO tooling.  Some of the components for the desktop system, like the Windows operating system, are not, but they are the exception.}\label{fig:mechatronic_tech_stack_solid_mechanics}
\end{figure}

The Aalto Solid Mechanics Laboratory's new control system provides a compelling case study on how control components may be selected when there are requirements placed on the control system beyond easy development.  It led to a need for high-speed IO and computation and a computer with multiple computation units with a CPU and an FPGA.\  Its tech stack is illustrated in figure~\ref{fig:mechatronic_tech_stack_solid_mechanics}.  The system as constructed has been able to meet the performance requirements and execute useful tests.  Development is on-going.

@@

This paper presented three case studies to demonstrate the utility of the novel theory of a mechatronic tech stack. These cases were diverse in their applications and methods and by viewing them through the lens of the tech stack, new knowledge of how mechatronic systems can be commissioned was gained. Also, by studying several stacks, I have formed a pool of tools to draw from. The mechatronic tech stack can be re-rendered to reflect this, as here in figure~\ref{fig:mechatronic_tech_stack_filled}.

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{assets/mechatronic_tech_stack_filled}
  \caption{The mechatronic tech stack for all of the case studies in one, with tools for each project highlighted in different colors.  Not only does the mechatronic tech stack present layers of abstraction, it also provides an idea of what tools to choose, similar to the web stack in figure~\ref{fig:web_stack}.}\label{fig:mechatronic_tech_stack_filled}
\end{figure}

@@

This thesis has also set forth a trajectory for future work in this topic, including expanding the scope of the tech stack from only control components to also hardware components, modeling the quality of tooling, and studying related topics like open-source software.  All this will be done in the author's doctoral studies.
